---
title: "Introduction to tvtools: Time-Varying Exposures for Survival Analysis"
author: "Timothy P. Copeland"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to tvtools: Time-Varying Exposures for Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The **tvtools** package provides comprehensive tools for creating time-varying exposure variables for survival analysis. This vignette introduces the core concepts and demonstrates the main functionality of `tvexpose()`.

## Why Time-Varying Exposures?

In many epidemiological and clinical studies, exposure status changes over time. For example:

- Patients may start or stop medications during follow-up
- Workers may change job duties and exposure levels
- Environmental exposures may vary over calendar time
- Participants may switch between different treatment regimens

Analyzing these situations requires **time-varying covariates** in survival models. Simply using baseline exposure status can lead to:

1. **Immortal time bias** - misclassifying pre-exposure time as exposed
2. **Time-varying confounding** - failing to account for exposure changes
3. **Loss of power** - ignoring information about exposure dynamics

## The Basic Concept

`tvexpose()` transforms period-based exposure data into a time-varying format suitable for survival analysis:

**Input:**
- Cohort dataset: One row per person with entry/exit dates
- Exposure dataset: Multiple rows per person, each representing an exposure period

**Output:**
- Time-varying dataset: Multiple rows per person-period, with exposure status for each interval

The output can be directly used with `survival::coxph()` and related functions.

## Data Structure Requirements

### Cohort Dataset

Your master cohort should have:
- Person identifier
- Study entry date (when follow-up begins)
- Study exit date (when follow-up ends - censoring or event)
- Baseline covariates
- Event indicator (optional)

```{r cohort_example, eval=FALSE}
cohort <- data.frame(
  id = 1:1000,
  study_entry = as.Date("2010-01-01"),
  study_exit = as.Date("2020-12-31"),
  age = rnorm(1000, 50, 10),
  female = rbinom(1000, 1, 0.5),
  event = rbinom(1000, 1, 0.1)
)
```

### Exposure Dataset

Your exposure data should have:
- Person identifier (matching cohort)
- Exposure start date
- Exposure stop date
- Exposure type/category (what the person was exposed to)

```{r exposure_example, eval=FALSE}
exposures <- data.frame(
  id = rep(1:1000, each = 5),
  exp_start = as.Date("2010-01-01") + sample(0:3650, 5000, replace = TRUE),
  exp_stop = as.Date("2010-01-01") + sample(365:3650, 5000, replace = TRUE),
  drug_type = sample(1:3, 5000, replace = TRUE)
)
```

## Basic Usage: Simple Time-Varying Exposure

The simplest use of `tvexpose()` creates time periods where exposure status changes:

```{r basic_example, eval=FALSE}
library(tvtools)
library(survival)
library(dplyr)

# Create time-varying exposure
tv_data <- tvexpose(
  cohort = cohort,
  exposures = exposures,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "exp_start",
  exp_stop = "exp_stop",
  exp_type = "drug_type",
  reference = 0  # 0 = unexposed
)

# View the structure
head(tv_data)
#   id      start       stop drug_type age female event
# 1  1 2010-01-01 2010-06-15         0  52      1     0
# 2  1 2010-06-15 2011-03-20         1  52      1     0
# 3  1 2011-03-20 2012-01-10         0  52      1     0
# 4  1 2012-01-10 2013-05-30         2  52      1     0
# 5  1 2013-05-30 2020-12-31         0  52      1     0

# Run Cox regression
cox_model <- coxph(Surv(start, stop, event) ~ drug_type + age + female,
                   data = tv_data)
summary(cox_model)
```

Each row represents a time period during which exposure status is constant. When exposure changes, a new row is created.

## Exposure Definition Types

`tvexpose()` supports multiple ways to define time-varying exposure through the `definition` parameter:

### 1. Basic Time-Varying (default)

Records actual exposure status during each period. Categories reflect the exposure type from the original data.

```{r basic_tv, eval=FALSE}
tv_basic <- tvexpose(
  cohort = cohort,
  exposures = exposures,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "exp_start",
  exp_stop = "exp_stop",
  exp_type = "drug_type",
  reference = 0,
  definition = "basic"  # This is the default
)

# drug_type values: 0 = unexposed, 1-3 = different drug types
```

### 2. Ever-Treated

Creates a binary indicator that switches from 0 to 1 at first exposure and remains 1 forever. Essential for correcting **immortal time bias** in ever-vs-never analyses.

```{r evertreated, eval=FALSE}
tv_ever <- tvexpose(
  cohort = cohort,
  exposures = exposures,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "exp_start",
  exp_stop = "exp_stop",
  exp_type = "drug_type",
  reference = 0,
  definition = "evertreated"
)

# Output variable: 0 = never treated (so far), 1 = ever treated
# Once switched to 1, stays 1 for all remaining follow-up
```

**Use case:** When you want to compare "ever users" vs "never users" while properly accounting for when treatment started.

### 3. Current/Former

Creates a three-level variable distinguishing never, current, and former exposure:

```{r currentformer, eval=FALSE}
tv_cf <- tvexpose(
  cohort = cohort,
  exposures = exposures,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "exp_start",
  exp_stop = "exp_stop",
  exp_type = "drug_type",
  reference = 0,
  definition = "currentformer"
)

# Output values:
# 0 = never exposed
# 1 = currently exposed
# 2 = formerly exposed (was exposed, but stopped)
# Note: Returns to 1 if person restarts after a gap
```

**Use case:** Distinguishing between active and past exposure effects (e.g., do protective effects persist after stopping?).

### 4. Duration Categories

Creates categories based on cumulative exposure duration:

```{r duration, eval=FALSE}
tv_duration <- tvexpose(
  cohort = cohort,
  exposures = exposures,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "exp_start",
  exp_stop = "exp_stop",
  exp_type = "drug_type",
  reference = 0,
  definition = "duration",
  duration_breaks = c(1, 5, 10),  # Break points in years
  duration_unit = "years"
)

# Output values:
# 0 = unexposed (0 years cumulative)
# 1 = < 1 year cumulative exposure
# 2 = 1 to < 5 years
# 3 = 5 to < 10 years
# 4 = 10+ years
```

**Use case:** Testing dose-response relationships based on cumulative duration.

### 5. Continuous Exposure

Creates a continuous variable tracking cumulative exposure:

```{r continuous, eval=FALSE}
tv_continuous <- tvexpose(
  cohort = cohort,
  exposures = exposures,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "exp_start",
  exp_stop = "exp_stop",
  exp_type = "drug_type",
  reference = 0,
  definition = "continuous",
  continuous_unit = "months"
)

# Output: Cumulative months of exposure at each time point
# Can be used as continuous predictor in Cox model
```

**Use case:** Modeling exposure as a continuous dose variable or for flexible spline models.

## Handling Data Quality Issues

Real-world exposure data often has gaps, overlaps, and other irregularities. `tvexpose()` provides several options to handle these:

### Grace Periods

Small gaps between prescriptions may represent refill delays rather than true cessation. Use `grace` to fill gaps:

```{r grace, eval=FALSE}
tv_grace <- tvexpose(
  cohort = cohort,
  exposures = exposures,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "exp_start",
  exp_stop = "exp_stop",
  exp_type = "drug_type",
  reference = 0,
  definition = "currentformer",
  grace = 30  # Fill gaps of 30 days or less
)

# Gaps ≤ 30 days are filled, treating brief interruptions as continuous exposure
```

You can also specify different grace periods for different exposure types:

```{r grace_bytype, eval=FALSE}
tv_grace_type <- tvexpose(
  cohort = cohort,
  exposures = exposures,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "exp_start",
  exp_stop = "exp_stop",
  exp_type = "drug_type",
  reference = 0,
  grace_by_type = list("1" = 30, "2" = 60, "3" = 90)
  # Drug type 1: 30-day grace, type 2: 60 days, type 3: 90 days
)
```

### Lag and Washout Periods

Model biological delays in exposure effects:

```{r lag_washout, eval=FALSE}
tv_lag <- tvexpose(
  cohort = cohort,
  exposures = exposures,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "exp_start",
  exp_stop = "exp_stop",
  exp_type = "drug_type",
  reference = 0,
  lag = 30,      # Exposure becomes active 30 days after start
  washout = 90   # Exposure effects persist 90 days after stop
)

# Exposure status changes 30 days AFTER exp_start
# Exposure status remains active until 90 days AFTER exp_stop
```

**Use cases:**
- `lag`: When drug effects don't begin immediately (e.g., slow-acting medications)
- `washout`: When drug effects persist after stopping (e.g., long half-life medications)

### Handling Overlapping Exposures

When exposure periods overlap (e.g., switching between medications), you can control how overlaps are resolved:

```{r priority, eval=FALSE}
# Give priority to certain exposure types
tv_priority <- tvexpose(
  cohort = cohort,
  exposures = exposures,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "exp_start",
  exp_stop = "exp_stop",
  exp_type = "drug_type",
  reference = 0,
  priority = c(3, 2, 1)  # Type 3 > Type 2 > Type 1
)

# When periods overlap, higher priority exposure is recorded
```

## Recency of Exposure

Track time since last exposure with recency categories:

```{r recency, eval=FALSE}
tv_recency <- tvexpose(
  cohort = cohort,
  exposures = exposures,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "exp_start",
  exp_stop = "exp_stop",
  exp_type = "drug_type",
  reference = 0,
  definition = "recency",
  recency_breaks = c(1, 5)  # Break points in years
)

# Output values:
# Current exposure
# < 1 year since last exposure
# 1 to < 5 years since last
# ≥ 5 years since last
```

**Use case:** Studying how quickly effects dissipate after stopping (washout dynamics).

## Keeping Additional Covariates

Bring baseline covariates into the time-varying dataset:

```{r keepvars, eval=FALSE}
tv_full <- tvexpose(
  cohort = cohort,
  exposures = exposures,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "exp_start",
  exp_stop = "exp_stop",
  exp_type = "drug_type",
  reference = 0,
  keep_vars = c("age", "female", "region", "comorbidity_score")
)

# Output includes all specified baseline variables
# Ready for regression without additional merging
```

## Diagnostics and Validation

Always check your time-varying data before analysis:

```{r diagnostics, eval=FALSE}
tv_checked <- tvexpose(
  cohort = cohort,
  exposures = exposures,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "exp_start",
  exp_stop = "exp_stop",
  exp_type = "drug_type",
  reference = 0,
  check_gaps = TRUE,      # Identify gaps in coverage
  check_overlaps = TRUE,  # Identify overlapping periods
  summarize = TRUE        # Display summary statistics
)

# Prints diagnostic information:
# - Number of persons with gaps
# - Number of overlapping periods
# - Exposure distribution summary
# - Person-time totals
```

## Complete Workflow Example

Here's a complete analysis workflow from data preparation through Cox regression:

```{r complete_workflow, eval=FALSE}
library(tvtools)
library(survival)
library(dplyr)

# 1. Load and prepare cohort data
cohort <- read.csv("my_cohort.csv") %>%
  mutate(
    study_entry = as.Date(study_entry),
    study_exit = as.Date(study_exit),
    event = as.integer(event_occurred)
  )

# 2. Load exposure data
exposures <- read.csv("drug_exposures.csv") %>%
  mutate(
    exp_start = as.Date(prescription_start),
    exp_stop = as.Date(prescription_end),
    drug_type = as.integer(drug_class)
  )

# 3. Create time-varying exposure with diagnostics
tv_data <- tvexpose(
  cohort = cohort,
  exposures = exposures,
  id = "patient_id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "exp_start",
  exp_stop = "exp_stop",
  exp_type = "drug_type",
  reference = 0,
  definition = "currentformer",
  grace = 30,  # 30-day grace period for refill gaps
  keep_vars = c("age", "sex", "comorbidity_index"),
  check_gaps = TRUE,
  summarize = TRUE
)

# 4. Verify data structure
head(tv_data, 20)
table(tv_data$drug_type_tv, useNA = "always")

# 5. Run survival analysis
cox_model <- coxph(
  Surv(start, stop, event) ~ drug_type_tv + age + sex + comorbidity_index,
  data = tv_data
)

summary(cox_model)

# 6. Check proportional hazards assumption
cox.zph(cox_model)

# 7. Visualize survival curves by exposure status
library(survminer)
ggsurvplot(
  survfit(cox_model, newdata = data.frame(
    drug_type_tv = c(0, 1, 2),
    age = 50,
    sex = 0,
    comorbidity_index = 2
  )),
  data = tv_data,
  legend.labs = c("Never", "Current", "Former"),
  xlab = "Time (years)",
  ylab = "Survival Probability"
)
```

## Common Use Cases

### 1. Pharmacoepidemiology: Drug Safety Study

```{r pharma_example, eval=FALSE}
# Evaluate cardiovascular risk associated with NSAIDs
# accounting for treatment switching and adherence gaps

tv_nsaid <- tvexpose(
  cohort = cohort_cvd,
  exposures = nsaid_prescriptions,
  id = "id",
  entry = "cohort_entry",
  exit = "cvd_date_or_censor",
  exp_start = "rx_start",
  exp_stop = "rx_stop",
  exp_type = "nsaid_class",
  reference = 0,
  definition = "currentformer",
  grace = 60,  # Allow 60-day gap (2 prescription cycles)
  lag = 14,    # Effects begin 14 days after start
  washout = 30 # Effects persist 30 days after stop
)

# Analyze with time-varying confounders
cox_nsaid <- coxph(
  Surv(start, stop, cvd_event) ~ nsaid_status + age + sex +
    diabetes + hypertension + prior_cvd,
  data = tv_nsaid
)
```

### 2. Occupational Cohort: Exposure-Response

```{r occupational_example, eval=FALSE}
# Study cancer risk by cumulative occupational exposure

tv_occupational <- tvexpose(
  cohort = worker_cohort,
  exposures = job_history,
  id = "worker_id",
  entry = "hire_date",
  exit = "exit_date",
  exp_start = "job_start",
  exp_stop = "job_stop",
  exp_type = "exposure_level",
  reference = 0,
  definition = "duration",
  duration_breaks = c(5, 10, 20),  # 0, <5, 5-10, 10-20, 20+ years
  duration_unit = "years"
)

# Test for dose-response trend
cox_occupational <- coxph(
  Surv(start, stop, cancer) ~ duration_category + age + smoking,
  data = tv_occupational
)

# Test linear trend
cox_trend <- coxph(
  Surv(start, stop, cancer) ~ as.numeric(duration_category) + age + smoking,
  data = tv_occupational
)
```

### 3. Clinical Trial: Treatment Switching

```{r clinical_example, eval=FALSE}
# Handle treatment switching in clinical trial
# Patients may switch from control to active treatment

tv_trial <- tvexpose(
  cohort = trial_participants,
  exposures = treatment_periods,
  id = "subject_id",
  entry = "randomization_date",
  exit = "last_contact",
  exp_start = "treatment_start",
  exp_stop = "treatment_stop",
  exp_type = "treatment_arm",
  reference = 0,  # 0 = control
  definition = "basic",
  track_switching = TRUE  # Creates switching indicator
)

# Analyze as-treated (accounting for crossover)
cox_astreated <- coxph(
  Surv(start, stop, event) ~ treatment_arm + baseline_severity,
  data = tv_trial
)

# Compare to intention-to-treat
cox_itt <- coxph(
  Surv(randomization_date, last_contact, event) ~
    randomized_arm + baseline_severity,
  data = trial_participants
)
```

## Performance Tips

For large datasets (>100K persons or >1M exposure periods):

1. **Filter unnecessary records early**: Remove exposure periods outside the study window before calling `tvexpose()`

2. **Use appropriate exposure definitions**: `"evertreated"` is faster than `"duration"` for simple analyses

3. **Limit diagnostics**: Set `check_gaps = FALSE` and `check_overlaps = FALSE` for production runs

4. **Consider data.table**: The package uses `dplyr` internally but can benefit from `data.table` for very large datasets

## Next Steps

- See `vignette("tvmerge-guide")` for merging multiple time-varying exposures
- Read function documentation: `?tvexpose` and `?tvmerge`
- Explore advanced options for complex exposure patterns
- Check package website for additional examples and tutorials

## References

1. Fisher LD, Lin DY. Time-dependent covariates in the Cox proportional-hazards regression model. *Annu Rev Public Health*. 1999;20:145-157.

2. Suissa S. Immortal time bias in pharmacoepidemiology. *Am J Epidemiol*. 2008;167(4):492-499.

3. Hernán MA, Robins JM. Using Big Data to Emulate a Target Trial When a Randomized Trial Is Not Available. *Am J Epidemiol*. 2016;183(8):758-764.

4. Therneau TM, Grambsch PM. *Modeling Survival Data: Extending the Cox Model*. Springer, 2000.

## Session Info

```{r session_info}
sessionInfo()
```
