---
title: "Creating Time-Varying Exposures with tvexpose"
author: "Timothy P. Copeland"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Time-Varying Exposures with tvexpose}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette demonstrates how to create time-varying exposure variables for survival analysis using `tvexpose()`. Time-varying exposures are essential when studying how exposures that change over time affect health outcomes.

## Why Time-Varying Exposures?

In many epidemiological studies, exposures are not static:

- **Pharmacoepidemiology**: Patients start, stop, and switch medications over time
- **Occupational studies**: Workers change jobs or exposure levels
- **Environmental epidemiology**: Pollutant levels fluctuate
- **Clinical research**: Treatment protocols change during follow-up

Standard survival analysis with baseline-only exposures can lead to:

1. **Immortal time bias**: Time before first exposure is incorrectly attributed to the exposed group
2. **Misclassification**: Exposure status at baseline may not reflect true exposure during follow-up
3. **Loss of information**: Changes in exposure status are ignored

`tvexpose()` creates properly formatted time-varying datasets that avoid these biases.

## The Challenge

Consider a simple example of HRT (hormone replacement therapy) exposure:

**Master Cohort:**
```
ID  Study_Entry  Study_Exit
1   2010-01-01   2020-12-31
```

**HRT Exposure Data:**
```
ID  Start       Stop        HRT_Type
1   2012-03-15  2014-06-30  1
1   2016-01-01  2018-12-31  2
```

We need to create a time-varying dataset that shows:
```
ID  Start       Stop        HRT_Status
1   2010-01-01  2012-03-14  0 (unexposed)
1   2012-03-15  2014-06-30  1 (Estrogen only)
1   2014-07-01  2015-12-31  0 (unexposed)
1   2016-01-01  2018-12-31  2 (E+P)
1   2019-01-01  2020-12-31  0 (unexposed)
```

`tvexpose()` automates this process, handling complex scenarios like overlapping exposures, gaps, and multiple exposure types.

## Loading Example Data

The tvtools package includes example datasets based on a simulated MS (multiple sclerosis) cohort:

```{r load_data}
library(tvtools)

# Load the example datasets
data(cohort)        # Master cohort: 1,000 persons
data(hrt_exposure)  # HRT exposure periods: 791 records for 391 persons
data(dmt_exposure)  # DMT exposure periods: 1,905 records for 758 persons

# Examine the cohort structure
str(cohort)
head(cohort)

# Examine HRT exposure data
str(hrt_exposure)
head(hrt_exposure)

# Examine DMT exposure data
str(dmt_exposure)
head(dmt_exposure)
```

### Dataset Descriptions

**cohort** (1,000 rows):
- `id`: Person identifier
- `study_entry`: Study entry date
- `study_exit`: Study exit date
- `age`: Age at baseline
- `female`: Sex (0=male, 1=female)
- `mstype`: MS type (1=RRMS, 2=SPMS, 3=PPMS)
- `edss_baseline`: Baseline EDSS disability score
- `region`: Geographic region

**hrt_exposure** (791 rows):
- `id`: Person identifier
- `rx_start`: HRT prescription start date
- `rx_stop`: HRT prescription stop date
- `hrt_type`: HRT type (1=Estrogen only, 2=E+P, 3=Other)
- `dose`: Daily dose in mg

**dmt_exposure** (1,905 rows):
- `id`: Person identifier
- `dmt_start`: DMT treatment start date
- `dmt_stop`: DMT treatment stop date
- `dmt`: DMT type (1=Interferon beta, 2=Glatiramer acetate, 3=Natalizumab, 4=Fingolimod, 5=Dimethyl fumarate, 6=Ocrelizumab)

## Basic Usage

### Simple Time-Varying Categorical Exposure

The most basic use case creates a time-varying categorical exposure variable:

```{r basic_categorical, eval=FALSE}
library(tvtools)
library(dplyr)

# Create time-varying HRT exposure
tv_hrt <- tvexpose(
  master = cohort,
  exposure_data = hrt_exposure,
  id = "id",
  start = "rx_start",
  stop = "rx_stop",
  exposure = "hrt_type",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  generate = "hrt_status"
)

# Examine output
head(tv_hrt, 20)
#   id      start       stop  hrt_status
# 1  1 2011-10-17 2020-12-31     0       # Never exposed
# 2  4 2011-03-06 2013-11-06     0       # Pre-exposure
# 3  4 2013-11-07 2015-09-03     1       # Estrogen only
# 4  4 2015-09-04 2016-05-05     0       # Gap
# 5  4 2016-05-06 2017-07-20     3       # Other
# ...

# Check distribution of exposure categories
table(tv_hrt$hrt_status, useNA = "always")
```

### Creating DMT Time-Varying Exposure

Similarly, create time-varying DMT exposure:

```{r basic_dmt, eval=FALSE}
tv_dmt <- tvexpose(
  master = cohort,
  exposure_data = dmt_exposure,
  id = "id",
  start = "dmt_start",
  stop = "dmt_stop",
  exposure = "dmt",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  generate = "dmt_status"
)

head(tv_dmt, 20)
```

## Exposure Type Options

`tvexpose()` supports several ways to define time-varying exposure status.

### 1. Categorical (Default)

Keeps the original exposure categories, with unexposed periods set to the reference value:

```{r categorical, eval=FALSE}
tv_categorical <- tvexpose(
  master = cohort,
  exposure_data = hrt_exposure,
  id = "id",
  start = "rx_start",
  stop = "rx_stop",
  exposure = "hrt_type",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  generate = "hrt_type_tv"
)

# Output values: 0 (unexposed), 1 (Estrogen), 2 (E+P), 3 (Other)
table(tv_categorical$hrt_type_tv)
```

### 2. Ever-Treated (Binary)

Creates a binary variable that switches from 0 to 1 at first exposure and stays 1 forever. This is the proper way to handle "ever-treated" analyses and avoid immortal time bias:

```{r evertreated, eval=FALSE}
tv_ever <- tvexpose(
  master = cohort,
  exposure_data = hrt_exposure,
  id = "id",
  start = "rx_start",
  stop = "rx_stop",
  exposure = "hrt_type",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  evertreated = TRUE,
  generate = "ever_hrt"
)

# Output values: 0 (never exposed yet), 1 (ever exposed)
table(tv_ever$ever_hrt)

# Verify: once someone becomes "ever-treated", they stay that way
tv_ever %>%
  group_by(id) %>%
  summarise(
    n_periods = n(),
    switches_to_1 = sum(diff(ever_hrt) == 1),  # Should be 0 or 1
    switches_to_0 = sum(diff(ever_hrt) == -1)  # Should always be 0
  ) %>%
  filter(switches_to_0 > 0)  # Should return 0 rows
```

### 3. Current/Former

Creates a trichotomous variable: 0 = never exposed, 1 = currently exposed, 2 = formerly exposed. If re-exposed after a gap, switches back to 1:

```{r currentformer, eval=FALSE}
tv_cf <- tvexpose(
  master = cohort,
  exposure_data = dmt_exposure,
  id = "id",
  start = "dmt_start",
  stop = "dmt_stop",
  exposure = "dmt",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  currentformer = TRUE,
  generate = "dmt_cf"
)

# Output values: 0 (never), 1 (current), 2 (former)
table(tv_cf$dmt_cf)

# Example person with gaps between treatments
tv_cf %>%
  filter(id == 1) %>%
  select(id, start, stop, dmt_cf)
#   id      start       stop  dmt_cf
# 1  1 2011-10-17 2013-01-07     0   # Never exposed
# 2  1 2013-01-08 2013-09-15     1   # Currently on DMT
# 3  1 2013-09-16 2019-06-08     2   # Formerly exposed
# 4  1 2019-06-09 2020-03-25     1   # Currently on DMT again
# ...
```

### 4. Duration Categories

Creates categories based on cumulative time of exposure. Useful for dose-response analyses:

```{r duration, eval=FALSE}
tv_duration <- tvexpose(
  master = cohort,
  exposure_data = dmt_exposure,
  id = "id",
  start = "dmt_start",
  stop = "dmt_stop",
  exposure = "dmt",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  duration = c(1, 3, 5),       # Cutpoints in years
  continuousunit = "years",     # Unit for duration
  generate = "dmt_duration"
)

# Output values:
# 0 = unexposed
# 1 = <1 year cumulative exposure
# 2 = 1-3 years cumulative exposure
# 3 = 3-5 years cumulative exposure
# 4 = 5+ years cumulative exposure
table(tv_duration$dmt_duration)
```

### 5. Continuous Cumulative Exposure

Creates a continuous variable tracking cumulative exposure time:

```{r continuous, eval=FALSE}
tv_continuous <- tvexpose(
  master = cohort,
  exposure_data = dmt_exposure,
  id = "id",
  start = "dmt_start",
  stop = "dmt_stop",
  exposure = "dmt",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  continuousunit = "months",
  generate = "cumul_dmt_months"
)

# Output: continuous variable (0 to max months of exposure)
summary(tv_continuous$cumul_dmt_months)
```

### 6. Recency Categories

Creates categories based on time since last exposure. Useful when exposure effects diminish over time:

```{r recency, eval=FALSE}
tv_recency <- tvexpose(
  master = cohort,
  exposure_data = hrt_exposure,
  id = "id",
  start = "rx_start",
  stop = "rx_stop",
  exposure = "hrt_type",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  recency = c(365, 1095),   # Cutpoints in days (1 year, 3 years)
  generate = "hrt_recency"
)

# Output values:
# 0 = never exposed
# 1 = currently exposed
# 2 = <1 year since last exposure
# 3 = 1-3 years since last exposure
# 4 = 3+ years since last exposure
table(tv_recency$hrt_recency)
```

## Handling Exposure Gaps

### Grace Period

Small gaps between prescriptions (e.g., due to refill delays) can be bridged using the `grace` parameter:

```{r grace, eval=FALSE}
# Without grace period - small gaps create unexposed periods
tv_no_grace <- tvexpose(
  master = cohort,
  exposure_data = hrt_exposure,
  id = "id",
  start = "rx_start",
  stop = "rx_stop",
  exposure = "hrt_type",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  grace = 0  # Default
)

# With 30-day grace period - gaps ≤30 days are bridged
tv_with_grace <- tvexpose(
  master = cohort,
  exposure_data = hrt_exposure,
  id = "id",
  start = "rx_start",
  stop = "rx_stop",
  exposure = "hrt_type",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  grace = 30  # Bridge gaps up to 30 days
)

# Compare number of periods
cat("Without grace:", nrow(tv_no_grace), "periods\n")
cat("With grace:", nrow(tv_with_grace), "periods\n")
```

### Merging Close Periods

The `merge_periods` parameter combines exposure periods of the same type that are close together:

```{r merge_periods, eval=FALSE}
tv_merged <- tvexpose(
  master = cohort,
  exposure_data = dmt_exposure,
  id = "id",
  start = "dmt_start",
  stop = "dmt_stop",
  exposure = "dmt",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  merge_periods = 120  # Merge same-type periods within 120 days (default)
)
```

### Fill Gaps

Extend the last exposure period by a fixed number of days:

```{r fillgaps, eval=FALSE}
tv_filled <- tvexpose(
  master = cohort,
  exposure_data = hrt_exposure,
  id = "id",
  start = "rx_start",
  stop = "rx_stop",
  exposure = "hrt_type",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  fillgaps = 90  # Extend last exposure by 90 days
)
```

## Timing Adjustments

### Lag Period

Model delayed biological effects by shifting exposure start dates forward:

```{r lag, eval=FALSE}
# Exposure effect begins 30 days after recorded start
tv_lagged <- tvexpose(
  master = cohort,
  exposure_data = dmt_exposure,
  id = "id",
  start = "dmt_start",
  stop = "dmt_stop",
  exposure = "dmt",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  lag = 30,  # 30-day lag
  generate = "dmt_lagged"
)
```

### Washout Period

Model residual effects that persist after exposure ends:

```{r washout, eval=FALSE}
# Exposure effect persists 90 days after recorded stop
tv_washout <- tvexpose(
  master = cohort,
  exposure_data = dmt_exposure,
  id = "id",
  start = "dmt_start",
  stop = "dmt_stop",
  exposure = "dmt",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  washout = 90,  # 90-day washout
  generate = "dmt_washout"
)
```

## Keeping Additional Variables

Bring baseline covariates from the master dataset into the time-varying output:

```{r keepvars, eval=FALSE}
tv_with_covars <- tvexpose(
  master = cohort,
  exposure_data = dmt_exposure,
  id = "id",
  start = "dmt_start",
  stop = "dmt_stop",
  exposure = "dmt",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  generate = "dmt_status",
  keepvars = c("age", "female", "mstype", "edss_baseline", "region")
)

# Output includes baseline covariates
names(tv_with_covars)
# [1] "id" "start" "stop" "dmt_status" "age" "female" "mstype" "edss_baseline" "region"
```

## By-Type Analysis

Create separate time-varying variables for each exposure type using `bytype = TRUE`:

```{r bytype, eval=FALSE}
# Ever-treated by each HRT type
tv_bytype_ever <- tvexpose(
  master = cohort,
  exposure_data = hrt_exposure,
  id = "id",
  start = "rx_start",
  stop = "rx_stop",
  exposure = "hrt_type",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  evertreated = TRUE,
  bytype = TRUE,
  generate = "ever_hrt"
)

# Creates: ever_hrt1, ever_hrt2, ever_hrt3
names(tv_bytype_ever)

# Current/former by each DMT type
tv_bytype_cf <- tvexpose(
  master = cohort,
  exposure_data = dmt_exposure,
  id = "id",
  start = "dmt_start",
  stop = "dmt_stop",
  exposure = "dmt",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  currentformer = TRUE,
  bytype = TRUE,
  generate = "dmt_cf"
)

# Creates: dmt_cf1, dmt_cf2, ..., dmt_cf6
names(tv_bytype_cf)
```

## Complete Survival Analysis Workflow

Here's a comprehensive example from raw data through final analysis:

```{r complete_workflow, eval=FALSE}
library(tvtools)
library(survival)
library(dplyr)

# ============================================================================
# STEP 1: Load data
# ============================================================================

data(cohort)
data(dmt_exposure)

# Examine cohort
cat("Cohort size:", nrow(cohort), "persons\n")
cat("Study period:", as.character(min(cohort$study_entry)), "to",
    as.character(max(cohort$study_exit)), "\n")

# ============================================================================
# STEP 2: Create time-varying DMT exposure (current/former)
# ============================================================================

tv_dmt <- tvexpose(
  master = cohort,
  exposure_data = dmt_exposure,
  id = "id",
  start = "dmt_start",
  stop = "dmt_stop",
  exposure = "dmt",
  reference = 0,
  entry = "study_entry",
  exit = "study_exit",
  currentformer = TRUE,
  grace = 30,  # Bridge 30-day gaps
  generate = "dmt_status",
  keepvars = c("age", "female", "mstype", "edss_baseline")
)

# ============================================================================
# STEP 3: Verify the time-varying data
# ============================================================================

cat("\nTime-varying data summary:\n")
cat("Total periods:", nrow(tv_dmt), "\n")
cat("Unique persons:", n_distinct(tv_dmt$id), "\n")

# Distribution of DMT status
cat("\nDMT status distribution:\n")
table(tv_dmt$dmt_status, useNA = "always")
# 0 = never, 1 = current, 2 = former

# Calculate person-time by status
tv_dmt %>%
  mutate(person_days = as.numeric(stop - start) + 1) %>%
  group_by(dmt_status) %>%
  summarise(
    n_periods = n(),
    person_days = sum(person_days),
    person_years = person_days / 365.25,
    .groups = "drop"
  ) %>%
  mutate(pct_person_time = round(100 * person_years / sum(person_years), 1)) %>%
  print()

# ============================================================================
# STEP 4: Prepare for survival analysis
# ============================================================================

# Convert dates to numeric for Surv()
tv_analysis <- tv_dmt %>%
  mutate(
    # Convert to numeric days since study start
    t_start = as.numeric(start - min(start)),
    t_stop = as.numeric(stop - min(start)),
    # Simulate an outcome (for demonstration)
    # In real analysis, this would come from your outcome data
    event = sample(c(0, 0, 0, 0, 1), n(), replace = TRUE)
  )

# ============================================================================
# STEP 5: Fit Cox proportional hazards model
# ============================================================================

# Model with time-varying DMT status
cox_model <- coxph(
  Surv(t_start, t_stop, event) ~ factor(dmt_status) + age + factor(female) +
    factor(mstype) + edss_baseline,
  data = tv_analysis,
  id = id
)

cat("\n========================================\n")
cat("Cox Proportional Hazards Model Results\n")
cat("========================================\n\n")
summary(cox_model)

# ============================================================================
# STEP 6: Model diagnostics
# ============================================================================

# Test proportional hazards assumption
cat("\nProportional Hazards Test:\n")
ph_test <- cox.zph(cox_model)
print(ph_test)

# ============================================================================
# STEP 7: Hazard ratios with 95% CI
# ============================================================================

# Extract hazard ratios
hr <- exp(coef(cox_model))
ci <- exp(confint(cox_model))

cat("\nHazard Ratios (95% CI):\n")
data.frame(
  HR = round(hr, 3),
  Lower = round(ci[, 1], 3),
  Upper = round(ci[, 2], 3)
) %>% print()
```

## Comparing Different Exposure Definitions

Sensitivity analyses often compare results across different exposure definitions:

```{r sensitivity, eval=FALSE}
library(tvtools)
library(survival)
library(dplyr)

data(cohort)
data(dmt_exposure)

# ============================================================================
# Create multiple exposure definitions
# ============================================================================

# Definition 1: Ever-treated (binary)
tv_ever <- tvexpose(
  master = cohort, exposure_data = dmt_exposure,
  id = "id", start = "dmt_start", stop = "dmt_stop", exposure = "dmt",
  reference = 0, entry = "study_entry", exit = "study_exit",
  evertreated = TRUE, generate = "dmt_ever",
  keepvars = c("age", "female", "mstype", "edss_baseline")
)

# Definition 2: Current/Former (trichotomous)
tv_cf <- tvexpose(
  master = cohort, exposure_data = dmt_exposure,
  id = "id", start = "dmt_start", stop = "dmt_stop", exposure = "dmt",
  reference = 0, entry = "study_entry", exit = "study_exit",
  currentformer = TRUE, generate = "dmt_cf",
  keepvars = c("age", "female", "mstype", "edss_baseline")
)

# Definition 3: Duration categories (<1, 1-3, 3+ years)
tv_dur <- tvexpose(
  master = cohort, exposure_data = dmt_exposure,
  id = "id", start = "dmt_start", stop = "dmt_stop", exposure = "dmt",
  reference = 0, entry = "study_entry", exit = "study_exit",
  duration = c(1, 3), continuousunit = "years", generate = "dmt_dur",
  keepvars = c("age", "female", "mstype", "edss_baseline")
)

# Definition 4: Current/Former with 60-day grace period
tv_cf_grace <- tvexpose(
  master = cohort, exposure_data = dmt_exposure,
  id = "id", start = "dmt_start", stop = "dmt_stop", exposure = "dmt",
  reference = 0, entry = "study_entry", exit = "study_exit",
  currentformer = TRUE, grace = 60, generate = "dmt_cf",
  keepvars = c("age", "female", "mstype", "edss_baseline")
)

# ============================================================================
# Compare results (with simulated outcome)
# ============================================================================

# Add simulated outcome and prepare for analysis
prepare_for_cox <- function(data, exp_var) {
  data %>%
    mutate(
      t_start = as.numeric(start - min(start)),
      t_stop = as.numeric(stop - min(start)),
      event = sample(c(0, 0, 0, 0, 1), n(), replace = TRUE)
    )
}

# Run models
results <- list()

# Model 1: Ever-treated
tv1 <- prepare_for_cox(tv_ever, "dmt_ever")
m1 <- coxph(Surv(t_start, t_stop, event) ~ dmt_ever + age + female, data = tv1, id = id)
results$ever_treated <- c(HR = exp(coef(m1)["dmt_ever"]),
                          exp(confint(m1)["dmt_ever", ]))

# Model 2: Current/Former
tv2 <- prepare_for_cox(tv_cf, "dmt_cf")
m2 <- coxph(Surv(t_start, t_stop, event) ~ factor(dmt_cf) + age + female, data = tv2, id = id)

cat("\nSensitivity Analysis Results:\n")
cat("==============================\n")

cat("\n1. Ever-treated:\n")
print(summary(m1)$conf.int[1, ])

cat("\n2. Current/Former:\n")
print(summary(m2)$conf.int[1:2, ])
```

## Common Pitfalls and Solutions

### Pitfall 1: Wrong Variable Types

```{r pitfall1, eval=FALSE}
# WRONG: Dates as character strings not converted
cohort_bad <- cohort
cohort_bad$study_entry <- as.character(cohort_bad$study_entry)

# tvexpose will attempt to convert, but may fail or produce warnings
# Always ensure dates are Date or POSIXct objects

# CORRECT: Proper date types
cohort_good <- cohort
class(cohort_good$study_entry)  # Should be "Date"
```

### Pitfall 2: Duplicate IDs in Master

```{r pitfall2, eval=FALSE}
# WRONG: Duplicate person in master dataset
cohort_dup <- rbind(cohort[1, ], cohort[1, ])  # Duplicate ID 1

tvexpose(
  master = cohort_dup,  # This will ERROR
  exposure_data = hrt_exposure,
  id = "id", start = "rx_start", stop = "rx_stop", exposure = "hrt_type",
  reference = 0, entry = "study_entry", exit = "study_exit"
)
# Error: master dataset has 1 duplicate ID(s)

# CORRECT: Each person appears once in master
cohort_unique <- cohort %>% distinct(id, .keep_all = TRUE)
```

### Pitfall 3: Conflicting Exposure Options

```{r pitfall3, eval=FALSE}
# WRONG: Multiple exposure types specified
tvexpose(
  master = cohort,
  exposure_data = hrt_exposure,
  id = "id", start = "rx_start", stop = "rx_stop", exposure = "hrt_type",
  reference = 0, entry = "study_entry", exit = "study_exit",
  evertreated = TRUE,     # Can't use both!
  currentformer = TRUE    # Can't use both!
)
# Error: Only one exposure type can be specified at a time

# CORRECT: Choose one exposure type
tvexpose(
  master = cohort,
  exposure_data = hrt_exposure,
  id = "id", start = "rx_start", stop = "rx_stop", exposure = "hrt_type",
  reference = 0, entry = "study_entry", exit = "study_exit",
  evertreated = TRUE  # OR currentformer, duration, etc.
)
```

### Pitfall 4: Exit Before Entry

```{r pitfall4, eval=FALSE}
# WRONG: Person has exit date before entry date
cohort_bad <- cohort
cohort_bad$study_exit[1] <- as.Date("2009-01-01")  # Before entry

tvexpose(
  master = cohort_bad,
  exposure_data = hrt_exposure,
  id = "id", start = "rx_start", stop = "rx_stop", exposure = "hrt_type",
  reference = 0, entry = "study_entry", exit = "study_exit"
)
# Error: 1 persons have study_exit < study_entry

# CORRECT: Validate dates before calling tvexpose
cohort %>%
  filter(study_exit < study_entry) %>%
  nrow()  # Should be 0
```

### Pitfall 5: NA Values in Critical Variables

```{r pitfall5, eval=FALSE}
# WRONG: Missing values in exposure or date columns
hrt_bad <- hrt_exposure
hrt_bad$hrt_type[1] <- NA

tvexpose(
  master = cohort,
  exposure_data = hrt_bad,
  id = "id", start = "rx_start", stop = "rx_stop", exposure = "hrt_type",
  reference = 0, entry = "study_entry", exit = "study_exit"
)
# Error: exposure variable 'hrt_type' contains 1 NA value(s)

# CORRECT: Handle missing values before tvexpose
hrt_clean <- hrt_exposure %>%
  filter(!is.na(hrt_type))  # Or impute as appropriate
```

### Pitfall 6: ID Type Mismatch

```{r pitfall6, eval=FALSE}
# WRONG: ID is numeric in one dataset, character in another
cohort_char <- cohort %>% mutate(id = as.character(id))
hrt_numeric <- hrt_exposure  # id is still numeric

tvexpose(
  master = cohort_char,
  exposure_data = hrt_numeric,
  id = "id", start = "rx_start", stop = "rx_stop", exposure = "hrt_type",
  reference = 0, entry = "study_entry", exit = "study_exit"
)
# Error: ID variable 'id' has different types in master and exposure_data

# CORRECT: Ensure consistent ID types
hrt_char <- hrt_exposure %>% mutate(id = as.character(id))
```

## Performance Considerations

For large datasets:

```{r performance, eval=FALSE}
# 1. Pre-filter exposure data to persons in cohort
cohort_ids <- unique(cohort$id)
hrt_filtered <- hrt_exposure %>%
  filter(id %in% cohort_ids)

# 2. Remove exposure periods completely outside study window
hrt_filtered <- hrt_filtered %>%
  inner_join(cohort %>% select(id, study_entry, study_exit), by = "id") %>%
  filter(rx_stop >= study_entry, rx_start <= study_exit) %>%
  select(-study_entry, -study_exit)

# 3. Process in batches for very large datasets
process_batch <- function(cohort_batch, exposure_data) {
  batch_ids <- unique(cohort_batch$id)
  exp_batch <- exposure_data %>% filter(id %in% batch_ids)

  tvexpose(
    master = cohort_batch,
    exposure_data = exp_batch,
    id = "id", start = "rx_start", stop = "rx_stop", exposure = "hrt_type",
    reference = 0, entry = "study_entry", exit = "study_exit"
  )
}

# Split into batches of 10,000
batch_size <- 10000
cohort_batches <- split(cohort, ceiling(seq_len(nrow(cohort)) / batch_size))
results <- lapply(cohort_batches, process_batch, exposure_data = hrt_exposure)
tv_combined <- bind_rows(results)
```

## Summary

Key takeaways for using `tvexpose()`:

1. **Choose the right exposure type**:
   - `evertreated = TRUE`: Binary ever/never (avoids immortal time bias)
   - `currentformer = TRUE`: Distinguishes current from former exposure
   - `duration = c(...)`: Categories based on cumulative exposure time
   - `continuousunit = "..."`: Continuous cumulative exposure
   - `recency = c(...)`: Categories based on time since last exposure

2. **Handle gaps appropriately**:
   - Use `grace` to bridge short gaps (prescription refills)
   - Use `merge_periods` to combine close same-type periods
   - Use `fillgaps` to extend the last exposure

3. **Adjust timing when needed**:
   - Use `lag` for delayed biological effects
   - Use `washout` for residual effects after exposure ends

4. **Include covariates**: Use `keepvars` to bring baseline variables into the output

5. **Validate your data**: Ensure no duplicate IDs, no NA values, and proper date types

## Next Steps

After creating time-varying exposures with `tvexpose()`:

1. **Merge multiple exposures**: Use `tvmerge()` to combine multiple time-varying exposures
   - See `vignette("tvmerge-guide")`

2. **Run survival analysis**: Use the output directly with `coxph()` from the survival package

3. **Perform sensitivity analyses**: Compare results across different exposure definitions

## Session Info

```{r session_info}
sessionInfo()
```
