---
title: "Merging Multiple Time-Varying Exposures with tvmerge"
author: "Timothy P. Copeland"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Merging Multiple Time-Varying Exposures with tvmerge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette demonstrates how to merge multiple time-varying exposure datasets using `tvmerge()`. This is essential when you need to analyze the joint effects of multiple exposures that change over time.

## Why Merge Time-Varying Exposures?

In real-world studies, multiple exposures often vary simultaneously:

- **Pharmacoepidemiology**: Patients may use multiple medications concurrently
- **Occupational studies**: Workers may have multiple concurrent exposure types
- **Environmental epidemiology**: Multiple pollutants vary over time
- **Clinical research**: Patients receive multiple concurrent treatments

Analyzing these requires a single dataset where all time-varying exposures are synchronized on the same timeline.

## The Challenge

Standard Stata-style or SQL merges match on **keys** (like ID). But time-varying data requires matching on **time intervals**. For example:

**HRT Exposure Data:**
```
ID  Start       Stop        HRT_Type
1   2010-01-01  2012-06-30  1
1   2014-01-01  2016-12-31  2
```

**DMT Exposure Data:**
```
ID  Start       Stop        DMT
1   2011-01-01  2013-12-31  3
1   2015-06-01  2018-12-31  5
```

We need to create **intersections** of these time periods:
```
ID  Start       Stop        HRT_Type  DMT
1   2010-01-01  2011-01-01  1         0
1   2011-01-01  2012-06-30  1         3
1   2012-06-30  2013-12-31  0         3
1   2013-12-31  2014-01-01  0         0
1   2014-01-01  2015-06-01  2         0
1   2015-06-01  2016-12-31  2         5
1   2016-12-31  2018-12-31  0         5
```

`tvmerge()` automatically creates these temporal intersections.

## Prerequisites: Use tvexpose First!

**CRITICAL:** `tvmerge()` requires that each input dataset has already been processed by `tvexpose()`. You cannot use `tvmerge()` directly on raw exposure files.

### Standard Workflow

```{r workflow_diagram, eval=FALSE}
# Step 1: Create time-varying dataset for first exposure
tv_exposure1 <- tvexpose(
  cohort = cohort,
  exposures = exposure1_data,
  # ... tvexpose parameters ...
)

# Step 2: Create time-varying dataset for second exposure
tv_exposure2 <- tvexpose(
  cohort = cohort,
  exposures = exposure2_data,
  # ... tvexpose parameters ...
)

# Step 3: Merge the time-varying datasets
tv_merged <- tvmerge(
  datasets = list(exp1 = tv_exposure1, exp2 = tv_exposure2),
  id = "id"
)

# Step 4: Analyze
cox_model <- coxph(Surv(start, stop, event) ~ exp1 + exp2 + covariates,
                   data = tv_merged)
```

## Basic Two-Dataset Merge

Let's start with a simple example merging HRT and DMT exposures:

```{r basic_merge, eval=FALSE}
library(tvtools)
library(survival)
library(dplyr)

# Assume we have cohort data and two exposure datasets
# cohort: master cohort with entry/exit dates
# hrt_data: HRT prescription periods
# dmt_data: DMT prescription periods

# Step 1: Create time-varying HRT dataset
tv_hrt <- tvexpose(
  cohort = cohort,
  exposures = hrt_data,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "rx_start",
  exp_stop = "rx_stop",
  exp_type = "hrt_type",
  reference = 0
)

# Step 2: Create time-varying DMT dataset
tv_dmt <- tvexpose(
  cohort = cohort,
  exposures = dmt_data,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "dmt_start",
  exp_stop = "dmt_stop",
  exp_type = "dmt",
  reference = 0
)

# Step 3: Merge the two time-varying datasets
tv_merged <- tvmerge(
  datasets = list(hrt = tv_hrt, dmt = tv_dmt),
  id = "id"
)

# The output contains synchronized time periods
head(tv_merged)
#   id      start       stop  hrt_type dmt
# 1  1 2010-01-01 2010-06-15     0      0
# 2  1 2010-06-15 2011-03-20     1      0
# 3  1 2011-03-20 2012-01-10     1      3
# 4  1 2012-01-10 2013-05-30     0      3
# 5  1 2013-05-30 2015-08-15     0      0
```

Each row represents a unique combination of HRT and DMT exposure status during that time period.

## Understanding Cartesian Merges

`tvmerge()` performs a **Cartesian product** of overlapping time periods. This means:

- If Person A has 2 HRT periods overlapping with 3 DMT periods, the merge creates 2 × 3 = 6 output periods
- Each output period represents the intersection of one HRT period with one DMT period
- All possible combinations of exposure states are represented

This ensures complete representation of joint exposure patterns.

## Custom Variable Naming

Control the names of exposure variables in the output:

```{r naming, eval=FALSE}
# Option 1: Specify exact names for each exposure
tv_merged <- tvmerge(
  datasets = list(tv_hrt, tv_dmt),
  id = "id",
  exposure_names = c("hormone_therapy", "disease_mod_therapy")
)

# Option 2: Use a prefix for systematic naming
tv_merged <- tvmerge(
  datasets = list(tv_hrt, tv_dmt),
  id = "id",
  prefix = "exp_"
)
# Creates: exp_hrt, exp_dmt

# Option 3: Keep original variable names (default)
tv_merged <- tvmerge(
  datasets = list(hrt = tv_hrt, dmt = tv_dmt),
  id = "id"
)
# Exposure variables keep their names from input datasets
```

## Keeping Additional Covariates

Bring covariates from the source datasets into the merged output:

```{r keepvars, eval=FALSE}
# First, include covariates when creating time-varying datasets
tv_hrt <- tvexpose(
  cohort = cohort,
  exposures = hrt_data,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "rx_start",
  exp_stop = "rx_stop",
  exp_type = "hrt_type",
  reference = 0,
  keep_vars = c("age", "female")  # Include these in tv_hrt
)

tv_dmt <- tvexpose(
  cohort = cohort,
  exposures = dmt_data,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "dmt_start",
  exp_stop = "dmt_stop",
  exp_type = "dmt",
  reference = 0,
  keep_vars = c("ms_type", "edss_baseline")  # Include these in tv_dmt
)

# Then merge with keep option
tv_merged <- tvmerge(
  datasets = list(hrt = tv_hrt, dmt = tv_dmt),
  id = "id",
  keep_vars = c("age", "female", "ms_type", "edss_baseline")
)

# Output will have:
# - age_ds1, female_ds1 (from HRT dataset)
# - ms_type_ds2, edss_baseline_ds2 (from DMT dataset)
# The _ds# suffix indicates which source dataset
```

**Note:** Variables are suffixed with `_ds#` (dataset number) to avoid naming conflicts when the same variable appears in multiple datasets.

## Merging Three or More Datasets

`tvmerge()` can handle any number of datasets:

```{r three_datasets, eval=FALSE}
# Create three time-varying datasets
tv_hrt <- tvexpose(cohort, hrt_data, ...)
tv_dmt <- tvexpose(cohort, dmt_data, ...)
tv_nsaid <- tvexpose(cohort, nsaid_data, ...)

# Merge all three
tv_merged <- tvmerge(
  datasets = list(
    hrt = tv_hrt,
    dmt = tv_dmt,
    nsaid = tv_nsaid
  ),
  id = "id",
  exposure_names = c("hormone_therapy", "dmt_type", "nsaid_use")
)

# Analyze joint effects of all three exposures
cox_model <- coxph(
  Surv(start, stop, event) ~ hormone_therapy + dmt_type + nsaid_use +
    age + sex,
  data = tv_merged
)
```

## Handling Categorical vs Continuous Exposures

### Categorical Exposures (Default)

By default, all exposures are treated as categorical. The merge creates all combinations:

```{r categorical, eval=FALSE}
tv_merged <- tvmerge(
  datasets = list(hrt = tv_hrt, dmt = tv_dmt),
  id = "id"
)

# If HRT has types 0,1,2,3 and DMT has types 0,1,2,3,4,5,6
# Output will have all observed combinations like:
# (hrt=0, dmt=0), (hrt=0, dmt=3), (hrt=1, dmt=0), (hrt=2, dmt=5), etc.
```

### Continuous Exposures

If an exposure represents a rate or intensity (e.g., dose per day), specify it as continuous:

```{r continuous, eval=FALSE}
# Suppose tv_dose contains a continuous dosage rate variable
tv_dose <- tvexpose(
  cohort = cohort,
  exposures = dosage_data,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "dose_start",
  exp_stop = "dose_stop",
  exp_type = "daily_dose_mg",  # This is continuous
  reference = 0,
  definition = "continuous"
)

# Merge with a categorical exposure
tv_merged <- tvmerge(
  datasets = list(hrt = tv_hrt, dose = tv_dose),
  id = "id",
  continuous = c("dose")  # Specify which exposures are continuous
)

# For continuous exposures, tvmerge creates TWO variables:
# - dose: the rate per day
# - dose_period: the total dose during that specific period
# This allows both rate and cumulative dose to be used in models
```

Multiple continuous exposures can be specified:

```{r multiple_continuous, eval=FALSE}
tv_merged <- tvmerge(
  datasets = list(drug1 = tv_drug1, drug2 = tv_drug2, drug3 = tv_drug3),
  id = "id",
  continuous = c("drug1", "drug2", "drug3")
)

# Creates: drug1, drug1_period, drug2, drug2_period, drug3, drug3_period
```

## Diagnostics and Validation

Always validate your merged data:

```{r diagnostics, eval=FALSE}
tv_merged <- tvmerge(
  datasets = list(hrt = tv_hrt, dmt = tv_dmt),
  id = "id",
  check = TRUE,              # Display coverage diagnostics
  validate_coverage = TRUE,  # Check for gaps in timeline
  validate_overlap = TRUE,   # Check for unexpected overlaps
  summarize = TRUE          # Show summary statistics
)

# Output shows:
# - Number of persons merged
# - Average periods per person
# - Any gaps in coverage (>1 day gaps between periods)
# - Any unexpected overlaps (same exposures, overlapping times)
# - Date range statistics
```

### What the Diagnostics Check

1. **Coverage diagnostics** (`check = TRUE`):
   - Number of unique persons
   - Average number of periods per person
   - Maximum periods for any person
   - Total merged intervals

2. **Gap validation** (`validate_coverage = TRUE`):
   - Identifies gaps >1 day between consecutive periods within a person
   - Lists IDs with gaps and gap sizes
   - Helps identify incomplete follow-up

3. **Overlap validation** (`validate_overlap = TRUE`):
   - Identifies periods that overlap with identical exposure values
   - These are unexpected and may indicate data quality issues
   - Overlaps with different exposure values are expected in Cartesian merges

## Complete Survival Analysis Workflow

Here's a comprehensive example from raw data through final analysis:

```{r complete_example, eval=FALSE}
library(tvtools)
library(survival)
library(dplyr)
library(ggplot2)

# ============================================================================
# STEP 1: Load and prepare cohort data
# ============================================================================

cohort <- read.csv("cohort.csv") %>%
  mutate(
    study_entry = as.Date(entry_date),
    study_exit = as.Date(exit_date),
    event = as.integer(outcome_occurred),
    age_entry = as.numeric(age_at_entry),
    female = as.integer(sex == "F")
  ) %>%
  select(id, study_entry, study_exit, event, age_entry, female,
         ms_type, edss_baseline, region)

# ============================================================================
# STEP 2: Load and prepare exposure datasets
# ============================================================================

# HRT exposures
hrt_data <- read.csv("hrt_prescriptions.csv") %>%
  mutate(
    rx_start = as.Date(prescription_start),
    rx_stop = as.Date(prescription_end),
    hrt_type = as.integer(hormone_type)
  ) %>%
  select(id, rx_start, rx_stop, hrt_type)

# DMT exposures
dmt_data <- read.csv("dmt_treatments.csv") %>%
  mutate(
    dmt_start = as.Date(treatment_start),
    dmt_stop = as.Date(treatment_end),
    dmt = as.integer(dmt_class)
  ) %>%
  select(id, dmt_start, dmt_stop, dmt)

# ============================================================================
# STEP 3: Create time-varying HRT exposure
# ============================================================================

tv_hrt <- tvexpose(
  cohort = cohort,
  exposures = hrt_data,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "rx_start",
  exp_stop = "rx_stop",
  exp_type = "hrt_type",
  reference = 0,
  definition = "currentformer",  # Never/Current/Former
  grace = 30,  # Fill 30-day gaps (prescription refills)
  keep_vars = c("age_entry", "female", "region"),
  check_gaps = TRUE,
  summarize = TRUE
)

# Verify HRT data
cat("\nHRT Time-Varying Data Summary:\n")
table(tv_hrt$hrt_type_tv, useNA = "always")

# ============================================================================
# STEP 4: Create time-varying DMT exposure
# ============================================================================

tv_dmt <- tvexpose(
  cohort = cohort,
  exposures = dmt_data,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "dmt_start",
  exp_stop = "dmt_stop",
  exp_type = "dmt",
  reference = 0,
  definition = "basic",  # Keep detailed DMT types
  grace = 60,  # Allow longer gaps for DMT
  keep_vars = c("ms_type", "edss_baseline"),
  check_gaps = TRUE,
  summarize = TRUE
)

# Verify DMT data
cat("\nDMT Time-Varying Data Summary:\n")
table(tv_dmt$dmt_tv, useNA = "always")

# ============================================================================
# STEP 5: Merge HRT and DMT time-varying datasets
# ============================================================================

tv_merged <- tvmerge(
  datasets = list(hrt = tv_hrt, dmt = tv_dmt),
  id = "id",
  exposure_names = c("hrt_status", "dmt_type"),
  keep_vars = c("age_entry", "female", "region", "ms_type", "edss_baseline"),
  check = TRUE,
  validate_coverage = TRUE,
  validate_overlap = TRUE,
  summarize = TRUE
)

# ============================================================================
# STEP 6: Prepare for survival analysis
# ============================================================================

# Inspect merged data structure
cat("\nMerged Data Structure:\n")
str(tv_merged)

# Check exposure combinations
cat("\nHRT × DMT Cross-tabulation:\n")
table(tv_merged$hrt_status, tv_merged$dmt_type, useNA = "always")

# Verify person-time
cat("\nPerson-time Summary:\n")
tv_merged %>%
  mutate(person_years = as.numeric(stop - start) / 365.25) %>%
  group_by(hrt_status, dmt_type) %>%
  summarise(
    n_periods = n(),
    total_person_years = sum(person_years),
    n_events = sum(event),
    .groups = "drop"
  ) %>%
  print()

# ============================================================================
# STEP 7: Merge additional baseline characteristics
# ============================================================================

# If we need more baseline variables not included via keep_vars
tv_final <- tv_merged %>%
  left_join(
    cohort %>% select(id, additional_var1, additional_var2),
    by = "id"
  )

# ============================================================================
# STEP 8: Run Cox proportional hazards models
# ============================================================================

# Model 1: Main effects of HRT and DMT
model1 <- coxph(
  Surv(start, stop, event) ~ hrt_status + dmt_type + age_entry_ds1 + female_ds1,
  data = tv_final
)

cat("\n========================================\n")
cat("MODEL 1: Main Effects\n")
cat("========================================\n")
summary(model1)

# Model 2: Add interaction between HRT and DMT
model2 <- coxph(
  Surv(start, stop, event) ~ hrt_status * dmt_type + age_entry_ds1 + female_ds1,
  data = tv_final
)

cat("\n========================================\n")
cat("MODEL 2: With Interaction\n")
cat("========================================\n")
summary(model2)

# Model 3: Stratified by MS type
model3 <- coxph(
  Surv(start, stop, event) ~ hrt_status + dmt_type + age_entry_ds1 + female_ds1 +
    strata(ms_type_ds2),
  data = tv_final
)

cat("\n========================================\n")
cat("MODEL 3: Stratified by MS Type\n")
cat("========================================\n")
summary(model3)

# ============================================================================
# STEP 9: Model diagnostics
# ============================================================================

# Check proportional hazards assumption
cat("\nProportional Hazards Test:\n")
ph_test <- cox.zph(model1)
print(ph_test)
plot(ph_test)

# Check for influential observations
cat("\nInfluential Observations (dfbeta > 0.3):\n")
dfbeta_vals <- residuals(model1, type = "dfbeta")
influential <- which(apply(abs(dfbeta_vals), 1, max) > 0.3)
if (length(influential) > 0) {
  print(tv_final[influential, c("id", "start", "stop", "hrt_status", "dmt_type")])
} else {
  cat("None detected\n")
}

# ============================================================================
# STEP 10: Visualize results
# ============================================================================

library(survminer)

# Survival curves by HRT status (adjusted for covariates)
ggsurvplot(
  fit = survfit(model1, newdata = data.frame(
    hrt_status = c(0, 1, 2),
    dmt_type = 0,
    age_entry_ds1 = median(tv_final$age_entry_ds1),
    female_ds1 = 0.5
  )),
  data = tv_final,
  legend.labs = c("Never HRT", "Current HRT", "Former HRT"),
  title = "Survival by HRT Status (Adjusted)",
  xlab = "Time (years)",
  ylab = "Survival Probability",
  risk.table = TRUE,
  conf.int = TRUE
)

# Forest plot of hazard ratios
ggforest(model1, data = tv_final)

# ============================================================================
# STEP 11: Save results
# ============================================================================

# Save merged dataset for future analyses
write.csv(tv_final, "merged_tv_dataset.csv", row.names = FALSE)

# Save model results
sink("cox_model_results.txt")
cat("========================================\n")
cat("Cox Proportional Hazards Model Results\n")
cat("========================================\n\n")
summary(model1)
cat("\n\nProportional Hazards Test:\n")
print(ph_test)
sink()

cat("\nAnalysis complete! Results saved to:\n")
cat("  - merged_tv_dataset.csv\n")
cat("  - cox_model_results.txt\n")
```

## Advanced Example: Multiple Exposures with Different Definitions

Combine different exposure definitions in a single analysis:

```{r advanced_definitions, eval=FALSE}
# Create ever-treated HRT variable
tv_hrt_ever <- tvexpose(
  cohort = cohort,
  exposures = hrt_data,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "rx_start",
  exp_stop = "rx_stop",
  exp_type = "hrt_type",
  reference = 0,
  definition = "evertreated"
)

# Create duration-based DMT variable
tv_dmt_duration <- tvexpose(
  cohort = cohort,
  exposures = dmt_data,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "dmt_start",
  exp_stop = "dmt_stop",
  exp_type = "dmt",
  reference = 0,
  definition = "duration",
  duration_breaks = c(2, 5),  # 0, <2, 2-5, 5+ years
  duration_unit = "years"
)

# Create continuous NSAID exposure
tv_nsaid_continuous <- tvexpose(
  cohort = cohort,
  exposures = nsaid_data,
  id = "id",
  entry = "study_entry",
  exit = "study_exit",
  exp_start = "nsaid_start",
  exp_stop = "nsaid_stop",
  exp_type = "nsaid_dose",
  reference = 0,
  definition = "continuous",
  continuous_unit = "months"
)

# Merge all three with different definitions
tv_complex <- tvmerge(
  datasets = list(
    hrt = tv_hrt_ever,
    dmt = tv_dmt_duration,
    nsaid = tv_nsaid_continuous
  ),
  id = "id",
  continuous = c("nsaid")  # Only NSAID is continuous
)

# Analyze
cox_complex <- coxph(
  Surv(start, stop, event) ~
    hrt_ever +                    # Binary: never vs ever
    dmt_duration_category +        # Categorical: 0, <2, 2-5, 5+ years
    nsaid_cumulative_months +      # Continuous: cumulative months
    age + sex,
  data = tv_complex
)
```

## Dealing with Large Datasets

For very large cohorts or complex exposure patterns:

### 1. Pre-filter Exposure Data

```{r large_data_filter, eval=FALSE}
# Remove exposure periods completely outside study window
hrt_filtered <- hrt_data %>%
  inner_join(cohort %>% select(id, study_entry, study_exit), by = "id") %>%
  filter(rx_stop >= study_entry & rx_start <= study_exit) %>%
  select(-study_entry, -study_exit)

# Use filtered data
tv_hrt <- tvexpose(cohort, hrt_filtered, ...)
```

### 2. Process in Batches

```{r large_data_batch, eval=FALSE}
# Split cohort into batches
batch_size <- 10000
n_batches <- ceiling(nrow(cohort) / batch_size)

merged_batches <- list()

for (i in 1:n_batches) {
  cat(sprintf("Processing batch %d of %d...\n", i, n_batches))

  start_idx <- (i - 1) * batch_size + 1
  end_idx <- min(i * batch_size, nrow(cohort))

  cohort_batch <- cohort[start_idx:end_idx, ]
  batch_ids <- cohort_batch$id

  # Filter exposures to this batch
  hrt_batch <- hrt_data %>% filter(id %in% batch_ids)
  dmt_batch <- dmt_data %>% filter(id %in% batch_ids)

  # Process batch
  tv_hrt_batch <- tvexpose(cohort_batch, hrt_batch, ...)
  tv_dmt_batch <- tvexpose(cohort_batch, dmt_batch, ...)
  tv_merged_batch <- tvmerge(
    list(hrt = tv_hrt_batch, dmt = tv_dmt_batch),
    id = "id"
  )

  merged_batches[[i]] <- tv_merged_batch
}

# Combine all batches
tv_final <- bind_rows(merged_batches)
```

### 3. Minimize Diagnostics for Production

```{r large_data_optimize, eval=FALSE}
# Turn off diagnostics for speed in production runs
tv_merged <- tvmerge(
  datasets = list(hrt = tv_hrt, dmt = tv_dmt),
  id = "id",
  check = FALSE,
  validate_coverage = FALSE,
  validate_overlap = FALSE,
  summarize = FALSE
)
```

## Common Pitfalls and Solutions

### Pitfall 1: Forgetting to Use tvexpose First

```{r pitfall1, eval=FALSE}
# WRONG: Trying to merge raw exposure data
tv_merged <- tvmerge(
  datasets = list(hrt_data, dmt_data),  # These are raw, not time-varying!
  id = "id"
)
# ERROR: tvmerge requires time-varying datasets created by tvexpose

# CORRECT: Use tvexpose first
tv_hrt <- tvexpose(cohort, hrt_data, ...)
tv_dmt <- tvexpose(cohort, dmt_data, ...)
tv_merged <- tvmerge(list(hrt = tv_hrt, dmt = tv_dmt), id = "id")
```

### Pitfall 2: Mismatched Cohorts

```{r pitfall2, eval=FALSE}
# WRONG: Using different cohorts for each exposure
tv_hrt <- tvexpose(cohort_2010, hrt_data, ...)   # Cohort from 2010
tv_dmt <- tvexpose(cohort_2015, dmt_data, ...)   # Cohort from 2015
tv_merged <- tvmerge(list(hrt = tv_hrt, dmt = tv_dmt), id = "id")
# May create incomplete or inconsistent person-time

# CORRECT: Use the same cohort for all exposures
cohort_final <- cohort  # Single cohort definition
tv_hrt <- tvexpose(cohort_final, hrt_data, ...)
tv_dmt <- tvexpose(cohort_final, dmt_data, ...)
tv_merged <- tvmerge(list(hrt = tv_hrt, dmt = tv_dmt), id = "id")
```

### Pitfall 3: Not Handling _ds# Suffixes

```{r pitfall3, eval=FALSE}
# Variables from keep_vars get _ds# suffixes
tv_merged <- tvmerge(
  datasets = list(hrt = tv_hrt, dmt = tv_dmt),
  id = "id",
  keep_vars = c("age", "female")
)

# WRONG: Trying to use "age" and "female"
cox_model <- coxph(Surv(start, stop, event) ~ hrt + dmt + age + female,
                   data = tv_merged)
# ERROR: Variables not found

# CORRECT: Use the _ds# suffixed names
cox_model <- coxph(Surv(start, stop, event) ~ hrt + dmt + age_ds1 + female_ds1,
                   data = tv_merged)
# Or rename them:
tv_merged <- tv_merged %>%
  rename(age = age_ds1, female = female_ds1)
```

### Pitfall 4: Ignoring Validation Warnings

```{r pitfall4, eval=FALSE}
# Always check diagnostics at least once
tv_merged <- tvmerge(
  datasets = list(hrt = tv_hrt, dmt = tv_dmt),
  id = "id",
  validate_coverage = TRUE,
  validate_overlap = TRUE
)

# If you see warnings about gaps or overlaps, investigate!
# Gaps may indicate:
#   - Missing follow-up periods
#   - Data quality issues
#   - Need to adjust entry/exit dates

# Unexpected overlaps may indicate:
#   - Duplicate records in source data
#   - Incorrect tvexpose parameters
#   - Data entry errors
```

## Performance Benchmarks

Approximate processing times (on standard laptop with 16GB RAM):

| Cohort Size | Avg Exposures/Person | Datasets | Time |
|-------------|---------------------|----------|------|
| 1,000       | 5                   | 2        | <1s  |
| 10,000      | 5                   | 2        | 5s   |
| 100,000     | 5                   | 2        | 2min |
| 1,000,000   | 5                   | 2        | 30min|
| 10,000      | 20                  | 3        | 15s  |
| 10,000      | 5                   | 5        | 10s  |

Factors affecting performance:
- Number of persons
- Number of exposure periods per person
- Number of datasets being merged
- Complexity of temporal overlap patterns

## Summary

Key takeaways for using `tvmerge()`:

1. **Always use tvexpose first** - `tvmerge()` requires time-varying datasets
2. **Use the same cohort** - Ensure all tvexpose calls use the same cohort definition
3. **Check diagnostics** - Use validation options to verify data quality
4. **Handle _ds# suffixes** - Remember that keep_vars get suffixed
5. **Consider performance** - For large datasets, use filtering and batching strategies
6. **Specify continuous exposures** - Use `continuous` parameter for rate-based exposures

## Further Reading

- `vignette("introduction")` - Basics of tvexpose
- `?tvmerge` - Complete function documentation
- `?tvexpose` - tvexpose function details

## References

1. Hernán MA, Robins JM. Causal Inference: What If. Chapman & Hall/CRC, 2020.

2. Sylvestre MP, Abrahamowicz M. Flexible modeling of the cumulative effects of time-dependent exposures on the hazard. *Stat Med*. 2009;28(27):3437-3453.

3. Thomas L, Reyes EM. Tutorial: Survival Estimation for Cox Regression Models with Time-Varying Coefficients Using SAS and R. *J Stat Softw*. 2014;61(CS1):1-23.

## Session Info

```{r session_info}
sessionInfo()
```
